---
title: "JustGreen Tool - Report"
author: "JustGreen Tool - Colorado State University"
output:
  html_document:
      css: style.css
date: "Report generated on `r Sys.Date()`"
params:
  cityName: NA,
  censusTractData : NA
  cityData : NA 
---


```{r load_packages_functions, echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load(leaflet, dplyr, readr)
sf_use_s2(FALSE)
### function for control number of character after the decimal place 
sigfig <- function(vec, n=3){ 
### function to round values to N significant digits
# input:   vec       vector of numeric
#          n         integer is the required sigfig  
# output:  outvec    vector of numeric rounded to N sigfig
formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 

} 
```


```{r dataProcessing, echo=FALSE, message=FALSE, warning=FALSE}
city <- "Birmingham city"
ct <- read_csv("~/trueNAS/work/justGreen/data/products/healthMeasures/allCT_2023_morDemStroke_with10percentAdjust_svi.csv")
city

```

# JustGreen ToolReport 
## Green Spaces and Health Impact 
**Location:** [CITY NAME, STATE] 
**Report Date:** [DATE] 

## Your Green Space Overview
* **Current Vegetation Level (NDVI):** [0.XX] (Scale: 0.0-1.0, where 1.0 = maximum vegetation) 
* **Ranking:** [XXth percentile] compared to similar US cities 
* **Population:** [XXX,XXX] residents in your area 

## How Your Area Compares
[Interactive gauge chart showing current NDVI level vs. city average and national benchmarks]

## Health Benefits of Current Green Spaces 
Based on scientific research, the current level of vegetation in your area provides these estimated annual health benefits: 

### Lives Protected 
* **Premature Deaths Prevented:** [XX] deaths annually 
* **Stroke Cases Prevented:** [X.X] cases annually
* **Dementia Cases Prevented:** [X.X] cases annually 

## Potential for Improvement 
### Current vs. Target Scenarios 

| Scenario | Vegetation Level (NDVI) | Additional Lives Saved | Additional Health Benefits |
|---|---|---|---|
| Current Level | [0.XX] | Baseline | Current benefits |
| Moderate Increase (+10%) | [0.XX] | +[XX] lives/year | +[XX]% stroke prevention |


## Where to Add Green Spaces 
[Map showing priority areas for vegetation expansion based on population density and current vegetation gaps] 

## Environmental Justice Considerations 
### Green Space Access by Demographics 

| Population Group 22] | Average NDVI Access 22] | Health Disparity Gap 22] |
|---|---|---|
| SVI quintile 1  | [0.XX]  | [X] premature health 25] |
| SVI quintile 2  | [0.XX]  | [X] premature health 28] |
| SVI quintile 3  | [0.XX]  | [X] premature health 31] |
| SVI quintile 4  | [0.XX]  | ☑ [X] premature health 35] |

**Priority for Equity:** Areas with the greatest health needs and lowest green space access should be prioritized for vegetation interventions. 

## Policy Recommendations 
### For Community Leaders 
1.  **Tree Canopy Goal:** define and increase urban tree coverage 
2.  **Equity Focus:** Prioritize green infrastructure in underserved neighborhoods 
3.  **Health Integration:** Include health impact assessments in all green space planning 

### For Residents 
1.  **Community Gardens:** Support local initiatives to create neighborhood growing spaces 
2.  **Tree Planting:** Participate in municipal tree planting programs
3.  **Green Advocacy:** Advocate for green space inclusion in new developments 

### For Urban Planners 
1.  **Zoning Requirements:** Mandate minimum green space ratios in new developments
2.  **Green Corridors:** Connect existing parks and green spaces with tree-lined pathways 
3.  **Climate Resilience:** Use vegetation to address urban heat island effects 

---
## Data Sources & Methodology 
This report uses satellite-derived vegetation data (NDVI), U.S. Census demographics, and peer-reviewed health impact models from environmental epidemiology research. Health estimates are based on established exposure-response relationships from longitudinal cohort studies across multiple cities. 

* **Scientific Basis:** Meta-analysis of 8+ million individuals showing 4% mortality reduction per 0.1 NDVI increase 
* **Data Sources:** NASA Landsat, U.S. Census ACS, CDC Social Vulnerability Index, CDC mortality statistics 
* **Geographic Resolution:** Census tract level for SVI analysis, block group level for vegetation data 

---
Report generated by [JustGreen Tool - Colorado State University]. For questions or technical details, contact: david.rojas@colostate.edu 
